Data Specialization for a Staged Lambda Calculus

### 2014-11-07 paper/splitting.tex
explaining split one rules

Given a term in the source language \lang\ where terms of
stage~\bbone\ and stage~\bbtwo\ may be interleaved, \emph{(stage)
  splitting} separates the stage~\bbone\ and stage~\bbtwo\
subcomputations into two separate terms, each performing the
subcomputations corresponding to each stage.  More precisely, given a
\lang\ term~$e$, splitting yields two terms $e_1$ and $e_2$ in a
unstaged target language \langmono\ such that~$e_1$ performs all the
stage~\bbone\ subcomputations in~$e$ and $e_2$~performs all of the
stage~\bbtwo\ subcomputations. Splitting preserves correctness in the sense that evaluation of~$e$ in
the staged language produces the same result as evaluating~$e_1$ and 
then~$e_2$ with the result from~$e_1$.

In this section, we present a
splitting algorithm for the source language \lang, specifying
precisely the algorithm and its correctness condition.  
We begin by defining a masking relation that serves to extract stage~\bbone\ and
stage~\bbtwo\ components of \lang\ partial values into \langmono\ terms that the outputs of splitting must match. We then 
move on to the specification of the splitting algorithm itself.

@@

% Intuitively, we require that the terms obtained by splitting a term $e$, when
% evaluated, agree with $e$ when evaluated according to our dynamic semantics. In
% the case of stage~\bbone\ terms, our dynamic semantics produce partial values
% containing information at \emph{both} stages, which goes against the idea that
% splitting separates the subcomputations belonging to each stage.

%For example, the partial value below contains both a stage~\bbone\ value 3, and a stage~\bbtwo\
%computation $\letin \yhat {\mathtt{7+5}} \yhat$.

% such as in the example below:

%For example, consider the stage 1 term
%\begin{lstlisting}
%1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
%\end{lstlisting}
%which evaluates via $\redonesym$ to a 
%
%residual table and multi-stage tuple

%This partial value contains both a stage~\bbone\ value 3, and a stage~\bbtwo\
%computation $\letin \yhat {\mathtt{7+5}} \yhat$.

Recall that \lang\ partial values and residual tables together may
contain content at \emph{both} stages. For example the term:
\begin{lstlisting}
1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
\end{lstlisting}
evaluates via $\redonesym$ to:
\begin{lstlisting}
2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
\end{lstlisting}
Intuitively, splitting this term should generate an $e1$ that reduces to the stage 1 content of the partial value (3), and an $e2$ corresponding to the stage~\bbtwo\
computation $\letin \yhat {\mathtt{7+5}} \yhat$.
To precisely delineate the stage~\bbone\ and stage~\bbtwo\ components of the partial value, we define 
\emph{partial-value masking} (or simply {\em masking}) functions, written $\masko{{-}}$ and $\maskt{{-};{-}}$, 
which take a residual table and partial value as input and emit a term in \langmono\ that corresponds to only the content belonging to a single stage. \ref{fig:valMask} defines these
masking functions.
% written $\masko{{-}}$ and $\maskt{{-};{-}}$.

\emph{Stage~\bbone\ masking} of the partial value $v$ (written $\masko{v}$) produces a term in \langmono\ containing exactly
the stage~\bbone\ content of $v$. Since \next\ terms only contain stage~\bbtwo\
content, $\masko{v}$ replaces all $\next$ subterms of $v$ with \texttt{()}.  Since lambdas may represent multi-stage computations, stage~\bbone\ masking splits their bodies as general stage~\bbone\ terms (as described in \ref{sec:split-one}), and returns
the stage~\bbone\ component. $\masko{v}$ always produces a value in \langmono\ that has
the same structure as~$v$.

\emph{Stage~\bbtwo\ masking} of the partial value $v$ and residual context $\xi$ (written $\maskt{\xi;v}$) produces a term in \langmono\ containing
exactly the stage~\bbtwo\ content of $[\xi;v]$. Stage~\bbtwo\ masking 
requires access to the residual table $\xi$ since all stage~\bbtwo\
content in $v$ has been lifted out into $\xi$. Stage~\bbtwo\ masking discards
stage~\bbone\ content in $v$ by replacing base constants with~\texttt{()},
replacing $\next~\hat{y}$ with~$\hat{y}$, and replacing all $n$-ary term
constructors with an $n$-tuple of masked subterms. Like stage~\bbone\ splitting, stage~\bbtwo\
splititing masks functions by splitting their bodies as general stage~\bbone\ terms, and
returning the stage~\bbtwo\ component. After discarding stage~\bbone\ content from $v$, stage~\bbtwo\ reifies $\xi$ around the result,
producing a term in \langmono.

@@

In general, the

>>>>

Given a term in the source language \lang\ where terms of
stage~\bbone\ and stage~\bbtwo\ may be interleaved, \emph{(stage)
  splitting} separates the stage~\bbone\ and stage~\bbtwo\
subcomputations into two separate terms, each performing the
subcomputations corresponding to each stage.  More precisely, given a
\lang\ term~$e$, splitting yields two terms $e_1$ and $e_2$ in a
unstaged target language \langmono\ such that~$e_1$ performs all the
stage~\bbone\ subcomputations in~$e$ and $e_2$~performs all of the
stage~\bbtwo\ subcomputations.

Splitting preserves correctness in the sense that evaluation of~$e$ in
the staged language produces the same result as evaluating~$e_1$ and
then~$e_2$ with the result from~$e_1$.  In this section, we present a
splitting algorithm for the source language \lang\, specifying
precisely the algorithm and the correctness condition.  We start by
defining a masking relation that helps us to isolate stage~\bbone\ and
stage~\bbtwo\ components of a partial value and then move on to the
specification of the splitting algorithms.

@@

Intuitively, we require that the terms obtained by splitting a term $e$, when
evaluated, agree with $e$ when evaluated according to our dynamic semantics. In
the case of stage~\bbone\ terms, our dynamic semantics produce partial values
containing information at \emph{both} stages, which goes against the idea that
splitting separates the subcomputations belonging to each stage.

For example, consider the stage 1 term
\begin{lstlisting}
1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
\end{lstlisting}
which evaluates via $\redonesym$ to a residual table and multi-stage tuple
\begin{lstlisting}
2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
\end{lstlisting}
This partial value contains both a stage~\bbone\ value 3, and a stage~\bbtwo\
computation $\letin \yhat {\mathtt{7+5}} \yhat$.

To clearly delineate these parts of the partial value, we define 
\emph{partial-value masking}, or simply {\em masking}, functions at each stage,
which compute from a residual table and partial value the content (expressed as
a term in \langmono) belonging to each stage. \ref{fig:valMask} defines these
masking functions, written $\masko{{-}}$ and $\maskt{{-};{-}}$.

The goal of $\masko{v}$ is to produce a term in \langmono\ containing exactly
the stage~\bbone\ content of $v$. Since \next\ terms only contain stage~\bbtwo\
content, we replace all $\next$ subterms of $v$ with \texttt{()}.  For lambdas,
since these represent potentially-multistage computations, we split their bodies
as general stage~\bbone\ terms (as described in \ref{sec:split-one}), and return
the stage~\bbone\ component. $\masko{v}$ is always a value in \langmono, and has
the same structure as~$v$.

The goal of $\maskt{\xi;v}$ is to produce a term in \langmono\ containing
exactly the stage~\bbtwo\ content of $[\xi;v]$. Unlike in stage \bbone\ masking,
here we need access to the residual table $\xi$, because all stage~\bbtwo\
content in $v$ has already been lifted out into $\xi$. To discard all
stage~\bbone\ information in $v$, we replace base constants with~\texttt{()},
replace $\next~\hat{y}$ with~$\hat{y}$, and replace all $n$-ary term
constructors with an $n$-tuple of their subterms, themselves masked. As before,
we mask functions by splitting their bodies as general stage~\bbone\ terms, and
use the stage~\bbtwo\ component.

After performing this transformation on $v$, we reify $\xi$ around the result,
producing a term in \langmono.

@@

When $e$ is a terminal (variable or base constant), the combined result
is formed by tupling $e$ with a \texttt{()} precomputation, 
and the resumer is just \texttt{()}.
When $e$ is a non-terminal (except $\next$),
splitting descends into $e$, recursively splitting subterms
and generating combined terms $c_1,\ldots,c_n$.
The new combined term of $e$ then works by calling $c_1,\ldots,c_n$,
destucting them into variables $y_1,\ldots,y_n$ for immediate results 
and $z_1,\ldots,z_n$ for precomputations.
The immediate result of $e$ is then formed by replacing its subterms with $y_1,\ldots,y_n$, 
and the precomputations is formed from the tuple of $z_1,\ldots,z_n$.

In the These combined terms are each
